$p$-adic precision

### 2014-02-15 padicprec.tex
Working on Section 2

The last two decades have seen a rise in the popularity of $p$-adic methods in
computational algebra.  For example,
\begin{itemize}
\item Bostan et al. \cite{Bostan} used Newton sums for polynomials over $\Zp$ to compute composed products for polynomials over $\Fp$;
\item Gaudry et al. \cite{Gaudry} used $p$-adic lifting methods to generate genus 2 CM hyperelliptic curves;
\item Kedlaya \cite{Kedlaya} used $p$-adic cohomology to count points on hyperelliptic curves over finite fields;
\item Lercier and Survent \cite{Lercier} computed isogenies between elliptic curves using $p$-adic differential equations.
\end{itemize}
Like real numbers, most $p$-adic numbers cannot be represented exactly, but instead must be
stored with some finite precision.  In this paper we focus on methods for handling $p$-adic precision
that apply across many different algorithms.

@@

If $E$ is a $K$-vector space, possibly of infinite dimension, then an
\emph{ultrametric norm} on $E$ is a map $\Vert\cdot\Vert : E \to \R^+$ satisfying:
\begin{enumerate}[(i)]
\item $\Vert x\Vert = 0$ if and only if $x = 0$;
\item $\Vert \lambda x\Vert = |\lambda| \cdot \Vert x\Vert$;
\item $\Vert x+y\Vert \leq \max(\Vert x\Vert, \Vert y\Vert)$.
\end{enumerate}
A \emph{$K$-Banach space} is a complete normed $K$-vector space.

\begin{deftn} % Separated definitions?
Let $E$ be a $K$-Banach space.
A \emph{lattice} in $E$ is an open bounded sub-$\OK$-module of $E$.
\end{deftn}

@@

\begin{deftn}
Suppose $E$ is a $K$-Banach space and $I$ a set.
A family $(x_i)_{i \in I} \subset E$ is a \emph{Banach basis} for $E$ if
every element $x \in E$ can be written
\[
x = \sum_{i \in I} \alpha_i x_i
\]
for scalars $\alpha_i \in K$ with $\alpha_i \to 0$, and $\Vert x \Vert = \sup_{i \in I} \lvert \alpha_i \rvert$.
\end{deftn}
Note that if $E$ is finite dimensional then the condition $\alpha_i \to 0$ is vacuous.
% Lattice from Banach basis

@@

Suppose that $E$ is a $K$-Banach space with basis $(x_i)_{i \in I}$.  We say that an element
$x \in E$ is \emph{exact} if there is a finite subset $J \subseteq I$ and scalars $\alpha_j \in R$ with
\begin{equation} \label{eq:exact_elt}
x = \sum_{j \in J} \alpha_j x_j.
\end{equation}
The set of exact elements is dense in $E$.

An approximate element is a pair $(x, H)$ where $x \in E$ is an exact element
and $H$ is a lattice in $E$.  Such a pair represents an undetermined element of the coset
$x + H$.  We will frequently write $x + O(H)$ to emphasize the fact that $H$ represents
the uncertainty in the value of the approximate element.  In the special case that $E = K = \Qp$,
we recover the standard notation $a + O(p^n)$ for an approximate $p$-adic element.

Note that using an arbitrary lattice to represent the precision of an approximate element
can reduce precision loss when compared to storing the precision of each coefficient
$\alpha_i$ in \eqref{eq:exact_elt} separately.  Consider the function 
$f : \Qp^2 \to \Qp^2$ mapping $(x,y)$ to $(x+y, x-y)$ and the problem of computing
$f \circ f(a + O(p^n), b + O(p^m))$.

Two approaches are possible. The first one consists in doing the 
computation step-by-step: we first compute $(z,t) = f(x,y)$ and we 
compute then $f(z,t)$ which gives the result. Proceeding this way, we 
see that $z$ and $t$ are both known up to precision $O(p^{\min(n,m)})$ 
are consequently the same holds for the coordinates of the result.
The second approach consists in remarking that $f \circ f$ maps
$(x,y)$ to $(2x, 2y)$. Computing $f \circ f(x,y)$ using the latter
formula, it is clear that the result we get is known up to precision
$(O(p^n), O(p^m))$ --- and even $(O(p^{n+1}), O(p^{m+1}))$ if $p = 2$.
As a conclusion, the second approach yields a more accurate result.
In other words, computing step-by-step creates ``artificial'' losses 
of precision.

Let us recall first 
that a datum involving elements of a complete ultrametric field $K$ 
(\emph{e.g.} $p$-adic numbers or power series) cannot be stored entirely 
on a computed because they basically convey an infinite amount of 
informations. That is why, instead of this datum itself, we generally 
work with neighborhoods of it.

In what follows, we shall consider neighborhoods of the form $x + H$ 
where $x$ is some element lying in a $K$-Banach $E$ and $H$ is a lattice 
in $E$. A very basic example is the approximate $p$-adic number $1 + 
O(p^N)$ which corresponds to the neighborhood of $1$ in $\Z_p$, which is 
the set of all $p$-adic numbers congruent to $1$ modulo $p^N$. We 
nevertheless emphasize that our definition is quite flexible and allows 
many neighborhoods that we do not consider in general. For instance, an 
``admissible'' neighborhood around a monic polynomial $P_0 \in \Q_p[X]$ 
of degree $d$ may be defined by congruences of the form $P(a_i) \equiv 
P_0(a_i) \pmod {p^{n_i}}$ ($1 \leq i \leq d$) for some fixed elements 
$a_i \in \Q_p$ and relative integers $n_i$.

In the context of computer science, we will often use the notation $x + 
O(H)$ to refer to $x + H$.

@@

@@

\subsubsection*{A first stupid example}

Let

@@

\subsubsection*{Lattices and computers}

Although elements of $K$ are not representable in thein entirety on a 
computer, lattices in finite dimensional $K$-vector spaces are as soon 
as the elements of the quotient $B_K(1)/B_K(r)$ are for all $r \in 
(0,1)$. Indeed, given a lattice $L$ in $K^d$, there exists by definition 
two positive rational numbers $a$ and $b$ such that 
\begin{equation}
\label{eq:incllattice}
B_K(a)^d \subset L \subset B_K(b)^d.
\end{equation}
Moreover, a lattice $L$ satisfying \eqref{eq:incllattice} is uniquely 
determined by its image in the quotient $B_K(b)^d / B_K(a)^d \simeq 
(B_K(1)/B_K(r))^d$ with $r = \frac a b$. Therefore, the data of $L$ can 
be encoded by a $(d \times d)$ matrix with coefficients in $B_K(1)/
B_K(r)$. Thus, if elements in the latter ring are representable on a 
computer, so are the lattices $L$ for which Eq.~\eqref{eq:incllattice} 
holds. Finally, an arbitrary lattice $L$ in $K^d$ can be represented
by a triple $(a,b,M)$ where $a$ and $b$ satisfies \eqref{eq:incllattice}
and $M$ is a $(d \times d)$ matrix over $B_K(1)/B_K(\frac a b)$ 
representing the image of $L$ in $B_K(b)^d / B_K(a)^d$.

We furthermore remark that if $K = \Q_p$ (for some prime number $p$), 
the quotient $B_K(1)/B_K(r)$ is isomorphism to some $\Z/p^n\Z$ (with $n 
= \lfloor - \log_p r \rfloor$) are therefore is representable on a 
computer. Thus, the above discussion applies when $K = \Q_p$. Similarly 
it also applies when $K$ is a finite extension of $\Q_p$ or a field of
Laurent series with coefficients in a finite field.

>>>>

The last two decades have seen a rise in the popularity of $p$-adic methods in
computational algebra.  For example,
\begin{itemize}
\item Bostan et al. \cite{Bostan} used Newton sums for polynomials over $\Zp$ to compute composed products for polynomials over $\Fp$;
\item Gaudry et al. \cite{Gaudry} used $p$-adic lifting methods to generate genus 2 CM hyperelliptic curves;
\item Kedlaya \cite{Kedlaya} used $p$-adic cohomology to count points on hyperelliptic curves over finite fields;
\item Lercier and Survent \cite{Lercier} computed isogenies between elliptic curves over finite fields using $p$-adic differential equations.
\end{itemize}
Like real numbers, most $p$-adic numbers cannot be represented exactly, but instead must be
stored with some finite precision.  In this paper we focus on methods for handling $p$-adic precision
that apply across many different algorithms.

@@

If $E$ is a $K$-vector space, possibly of infinite dimension, then an
\emph{ultrametric norm} on $E$ is a map $\Vert\cdot\Vert : E \to \R^+$ satisfying:
\begin{enumerate}[(i)]
\item $\Vert x\Vert = 0$ if and only if $x = 0$;
\item $\Vert \lambda x\Vert = |\lambda| \cdot \Vert x\Vert$;
\item $\Vert x+y\Vert \leq \max(\Vert x\Vert, \Vert y\Vert)$.
\end{enumerate}
A \emph{$K$-Banach space} is a complete normed $K$-vector space.
A \emph{lattice} in a $K$-Banach space $E$ is an open bounded sub-$\OK$-module of $E$.

@@

Suppose $E$ is a $K$-Banach space and $I$ a set.
A family $(x_i)_{i \in I} \subset E$ is a \emph{Banach basis} for $E$ if
every element $x \in E$ can be written
\[
x = \sum_{i \in I} \alpha_i x_i
\]
for scalars $\alpha_i \in K$ with $\alpha_i \to 0$, and $\Vert x \Vert = \sup_{i \in I} \lvert \alpha_i \rvert$.
Note that if $E$ is finite dimensional then the condition $\alpha_i \to 0$ is vacuous.

Given a basis $(x_i)_{i \in I}$ and a sequence $(r_i)_{i \in I}$ with $r_i \in \R_{>0}$, the sets
\begin{align*}
B_E((x_i),(r_i)) &= \Big\{ \sum_{i \in I} \alpha_i x_i : \lvert \alpha_i \rvert \le r_i \Big\}, \\
B^-_E((x_i),(r_i)) &= \Big\{ \sum_{i \in I} \alpha_i x_i : \lvert \alpha_i \rvert < r_i \Big\}
\end{align*}
are lattices precisely when the $r_i$ are bounded.  If we have equipped $E$ with a distinguished basis
then we may drop $(x_i)$ from the notation for $B^{(-)}_E((x_i),(r_i))$.
% Is every lattice of this shape for some Banach basis?

\subsubsection*{Lattices and computers.}

Suppose that $E \simeq K^d$ is finite dimesional.  Then if $H \subset E$
is a lattice then there exist $a, b \in \Q_{>0}$ with
\begin{equation}
\label{eq:incllattice}
B_K(a)^d \subset H \subset B_K(b)^d.
\end{equation}
Set $r = \frac a b$ and $R_r = \OK/B_K(r)$.  
Then a lattice $H$ satisfying \eqref{eq:incllattice} is uniquely 
determined by its image in the quotient $B_K(b)^d / B_K(a)^d \simeq 
R_r^d$.  Since $R \cap \OK$ is dense in $\OK$, elements of $R_r$ may be
represented exactly.  Thus $H$ may be encoded as a 
$(d \times d)$ matrix with coefficients in $R_r$.  For example, when
$K = \Qp$ the ring $R_r$ is just $(\Z / p^n\Z)$ for $n = \lfloor - \log_p r \rfloor$.

@@

Suppose that $E$ is a $K$-Banach space with basis $(x_i)_{i \in I}$.
\begin{deftn}
\begin{itemize}
\item An element $x \in E$ is \emph{exact} if there is a finite subset $J \subseteq I$ and scalars $\alpha_j \in R$ with
\begin{equation} \label{eq:exact_elt}
x = \sum_{j \in J} \alpha_j x_j.
\end{equation}

\item An \emph{approximate element} is a pair $(x, H)$ where $x \in E$ is an exact element
and $H$ is a lattice in $E$.
\end{itemize}
\end{deftn}

The pair $(x, H)$ represents an undetermined element of the coset
$x + H$.  We will frequently write $x + O(H)$ to emphasize the fact that $H$ represents
the uncertainty in the value of the approximate element.  In the special case that $E = K = \Qp$,
we recover the standard notation $a + O(p^n)$ for an approximate $p$-adic element.
Note that the set of exact elements is dense in $E$, so every element of $E$ can be approximated.

This definition encapsulates the main practical suggestion of the paper with regards to representing
vector spaces, matrices, polynomials and power series over $K$: one should \textbf{separate} the
approximation from the precision.  In the rest of this section we discuss some of the benefits made
possible by such a separation.

Note first that using an arbitrary lattice to represent the precision of an approximate element
can reduce precision loss when compared to storing the precision of each coefficient
$\alpha_i$ in \eqref{eq:exact_elt} separately.  Consider the function 
$f : \Qp^2 \to \Qp^2$ mapping $(x,y)$ to $(x+y, x-y)$ and the problem of computing
$f \circ f(a + O(p^n), b + O(p^m))$.

Write $(e_1, e_2)$ for the standard basis of $E=\Qp^2$.  Since $f$ is linear, the image of
the approximation $\bigl((a,b), B_E\bigl((e_1,e_2),(p^{-n}, p^{-m})\bigr)\bigr)$ is
$\bigl((a+b, a-b), B_E\bigl((e_1+e_2, e_1-e_2), (p^{-n}, p^{-m})\bigr)\bigr)$.
For $p \ne 2$, applying $f$ again yields $\bigl((2a, 2b), B_E\bigl((e_1, e_2), (p^{-n}, p^{-m})\bigr)\bigr)$.
Storing precisions on each coordinate of a vector amounts to only allowing precisions of the form
$B_E((r_i))$ with respect to the standard basis.  One would be forced to truncate the intermediate precision
to $B_E(p^{-\min(m, n)})$, yielding an artificial loss of precision in the final result.

@@

In addition to allowing for a more flexible representation of the precision of an element,
the separation of precision from approximation has other benefits as well.  If the precision
is encoded with the approximation, certain algorithms become unusable because of their
numerical instability.  For example, the Karatsuba algorithm for polynomial multiplication

@@

@@
